Fused regression for multi-source gene regulatory network inference

### 2016-01-09 paper1.tex
zack committing

\title{Fused regression for multi-source network inference}
\author{Kari Y. Lam, Zachary M. Westrick, Christian L. M\"{u}ller, Lionel Christiaen, Richard Bonneau}
\maketitle

@@

We present two methods for network inference based on linear estimates of gene expression dynamics, extending existing linear dynamical-systems methods for network inference \cite{bonneau_predictive_2007, arrieta-ortiz_experimentally_2015}. The core of both methods is the observation that biological information about the relatedness of two networks can be used to constrain which network coefficients should be similar to one another in a multi-source network inference problem (ie orthologous TFs should regulate orthologous genes), and that these constraints can be efficiently represented as penalties in a least-squares regression problem. Our first method -- fused ridge -- uses an L2 penalty on the differences between a priori similar interactions, and is useful where the relationships between networks (similarity of genes between data sets) is reliable. In the case where both networks contain an identical set of genes and TFs, this approach can be thought of as parametrically interpolating between treating the data sources separately and combining them together. However, our method allows useful pooling of data even when the overlap between genes is incomplete. Our second method -- adaptive fusion -- uses a non-concave saturating fusion penalty to simultaneously learn the constrained networks and to learn which constraints should be relaxed (ie which parts of the network are genuinely different). With this approach, we seek to test the hypothesis of conservation of interactions across networks.

@@

We test the ability of fused L2 and adaptive fusion to improve network recovery on both synthetic data and Bacillus subtilis and Bacillus anthracis, showing the applicability of our method in combining different datasets, and leveraging similarity across organisms as well as within a network in order to improve network inference. We explore the circumstances under which each approach is optimal, and evaluate the robustness of adaptive fusion to incorrect orthology, simulating the biologically relevant case of neofunctionalization.

@@

%We can then define $W_{i,j} = \beta_{(j-1) \times N + i, (j-1) \times M + j}$ which is the desired form.
This is equivalent to the original problem due to the block structure of matrix multiplication. In an ordinary regression problem each response variable can be solved independently as a large number of small regression problems. However, in fused regression, we append additional rows to the design matrix $A$ that link entries of $W$ associated with different response variables. As a result, these linked response variables must be solved simultaneously through vectorization. Two response variables are linked by a fusion constraint if any of the regulatory weights affecting those genes are linked by a fusion constraint. Two response variables must be solved simultaneously if there is any chain of linked response variables connecting them. However, every other response variable can be solved separately. In biological terms, the regulators of two genes (whether in the same species, or different species) must be solved together if there is a fusion constraint linking those genes' regulators, or if there is a chain of such constraints. This is important because vectorization produces a design matrix with a number of rows and columns equal to the sum of the number of rows and columns of the design matrices associated with each of the response variables (genes) being solved for. If the networks for a large number of Genes are solved simultaneously, the system of equations can quickly become intractible to solve.

@@

\section{Results}
We tested the ability of fused regression to improve network inference performance, as well as the ability of adaptive-fusion to learn the conservation of gene interactions. We measured network inference performance with the area under the precision recall curve (AUPR).

\subsection{Using fused regression to learn related networks}
Using related data for network inference allows us to borrow statistical power from other sources, effectively increasing the sample size and boosting the sensitivity and specificity of learned interactions. We used synthetic data to approximate two related biological processes and show that simultaneously learning the networks describing them using fused L2 regression to share information about related interactions results in an improvement in our ability to recover the true networks. Adding conditions from a related dataset has an effect on network inference which is akin to adding experimental data, without doing additional experiments.

@@

As expected, when the conserved subgraph is very similar, increasing the weight of the fusion constraint $\lambda_S$ improved network recovery (Figure 2). Networks with higher orthology coverage saw a greater benefit from fusion.

When the conservation is weak, and even the portions of the network which are conserved have deviated from each other significantly, performance may not continue to improve by increasing the fusion weight; we simulate this by adding noise to the conserved subgraph (Figure 3). We show that even for networks where the conserved subgraph is weakly conserved, there exists a 'sweet spot' where fusion regression improves network recovery.

\subsection{Fused regression improves performance on both the constrained and non-constrained parts of the network}
Our method learns networks simultaneously, using predictions about conservation of TF-gene interactions to share information between sources. Our approach is useful for learning networks from similar sources such as different cell-lines from the same species, where there exists a one-to-one mapping of genes, as well as datasets where the networks are expected to be more different than similar, and there exists limited orthology mapping. For the latter case, we are interested in knowing if performance gains from fused regression are limited to those interactions which have fusion constraints. To test this, we used sets of 20 TF by 200 gene synthetic networks, with varying proportions of orthology information about TFs and genes. We varied lamS and computed AUPR on the portion of the network that had fusion constraints, the portion where TF-gene interactions from one network could be mapped to putative analogous interactions in the second network based off of orthology; the unconstrained subnetwork; and the whole network (Figure 5). As expected, performance improved as the fusion penalty weight increased. Interestingly, performance gains were seen even in the portion of the network that is unconstrained by fusion. By constraining part of an underdetermined system, we obtain gains in even the unconstrained part.

\subsection{Adaptive fusion successfully identifies and unfuses 'neofunctionalized' genes}
We recognize that orthology prediction is not a perfect proxy for functional conservation. We implemented an adaptive fusion algorithm that attempts to optimize a nonconvex saturating penalty function on differences between fused interactions (Figures 5-6). Pairs of interactions which are very dissimilar even after fusion, which sit in the flat portion of this penalty function, are effectively ``unfused,'' and no further penalty is incurred as differences in interaction weights grow. Because our network inference problem is underdetermined, meaning that multiple networks would fit our data equally well, the ``unfusing'' of constraints is convincing evidence for the rejection of a network which uses orthology to predict similarity in interaction weights. ''Unfusing'' suggests neofunctionalization of one or more of the genes involved in this pair of interactions.

We performed a simulation to assess the ability of our adaptive fusion algorithm to learn which parts of two input networks are conserved when orthology information is faulty, to mimic orthologs which are not functionally analogous  (Figure 7-8). We generated synthetic fused networks and introduced error in the fusion constraints by adding false positives and negatives to the orthology information given to the solver. Because we knew which entries in the orthology mapping were ``fake'' (not reflected in the generation of the networks), we could correctly label fusion constraints that involved one or more ``fake'' mappings. We verified that adaptive-fusion unfused mostly ``fake'' interactions, while leaving truly analogous interactions fused. We compared the recovery of interaction weights which were accurately fused, and recovery of interaction weights which were inaccurately fused due to incorrect orthology information. Because fused L2 heavily penalizes large differences between weights which are predicted to be similar (Figures 8-9), the error for inaccurately fused interactions is higher than solving without fusion. When we use our adaptive fusion approach, we preserve the gain in performance on accurately fused interactions, while reducing the error for weights of interactions involving "fake" orthologies.

\subsection{Cross-species network inference using bacterial data}
We used gene-expression data from B. subtilis and B. anthracis in order to assess performance gains of fused regression on real data. Our B subtilis data set consists of 360 time-series and steady-state observations of 4891 genes during development. Our B. anthracis dataset consists of 72 time-series and steady-state observations of 5536 genes comprising data from developmental and iron-starvation conditions. There were 247 known Transcription Factors (TFs) in the B. subtilis dataset, and 248 TFs in the B. anthracis dataset.

We obtained 1870 one-to-one orthologs from Inparanoid \cite{ostlund_inparanoid_2010}, which produced 177650 fusion-constraints between gene interactions within the two species. Although this number seems very large, it represents only $14.7\%$ of the regulatory interaction matrix in B. subtilis and $12.9\%$ in B. anthracis. The use of such dissimilar species is a test of whether fusion can improve network-inference even when the overlap in interactions is very small.

In order to evaluate network inference performance, and for use as priors, we used a gold standard of 2896 known B. subtilis interactions with corresponding activation and repression sign. Of these 2896 priors, 968 had corresponding interactions in B. anthracis. Based on our simulation results, we can expect the greatest gains in network-inference performance from fusion when the species of interest has a small number of available conditions, but data is abundant in a related species. However, in order to evaluate performance objectively a gold-standard of known interactions is necessary. As a result, we can only evaluate network recovery for B. subtilis, and B. subtilis also has the majority of our conditions. In order to simulate the data-poor regime, we subsampled our B. subtilis data. Specifically, we divided our B. subtilis data into $k$ folds, and then for each fold fit a network to the B. subtilis data from that fold alone fused to the entire 72 B. anthracis conditions. We can then compute performance metrics (ie AUPR) for each fold, and use their variability across folds to test for significance of any improvements relating to the use of fusion constraints.

When we applied fused L2 to our subsampled B. subtilis and B. anthracis datasets, we did not see a significant improvement over solving the networks separately. Inspection of precision recall curves shows that this improvement is a result of increased precision at low values of recall. Essentially, precision increases for the high confidence part of the network but not for the low confidence part of the network. As a result, the improvements in performance may be more useful than would be suggested by AUPR alone. Interactions suggested by genome wide analysis need to be experimentally verified, and the ranking of interactions produced by a network inference algorithms is a useful guide to what order these experiments should be carried out in. The high confidence part of the network represents interactions that are likely candidates for experimental validation, so improvement in recovery of these interactions is particularly useful in guiding research.

\subsection{Within-species fusion using similarity of promoter region}
Information about the similarity of TF-gene interactions can also come from knowledge about the promoter region; this may be a better predictor of target similarity than orthology. In bacteria, genes within the same operon are under the control of the same promoter. We predict, therefore, that genes within the same operon will be regulated similarly by the same transcription factors. We apply fusion regression by creating fusion constraints between a given transcription factor and genes within the same operon, and show a boost in B. subtilis network recovery using within-species fusion.

\subsection{Integrating datasets using fusion regression}
Although there are many large-scale collaborations which attempt to make protocols as uniform as possible for comparability between datasets generated by different labs, there still exists technical and biologial variability between many experiments attempting to capture the same or similar experimental condition. With the advent of new technology, such as RNAseq and single-cell RNAseq, microarray is no longer the dominant assay for genome-wide expression, but it is important to be able to combine datasets generated using different technologies, lest we lose valuable information. Currently, the most widely used approach to combining datasets for network inference is to learn networks from disparate datasets separately, then rank combine the networks as in Marbach et al \cite{marbach_revealing_2010}. We compare this approach to using fusion regression. We include, along with our B. subtilis dataset, a previously published dataset containing 269 samples covering 104 conditions, obtained using a different microarray and different strain of B. subtilis \cite{nicolas2012condition}. We compare performance when learning the networks separately, learning the networks separately followed by rank combining, and learning the networks simultaneously using fusion regression, and show improvement on network inference using our fused L2 approach. Importantly, our approach simultaneously learns multiple networks, rather than learning networks separately than combining into one.

\subsection{Transcription factor activity estimation integrates into fusion regression approach}
We test a combination of our fused regression approach with a method for estimating transcription factor activities (TFA). Rather than modeling gene expression using transcription factor mRNA abundance, we attempt to fit gene expression as a function of transcription factor activity. We estimate TFA based on known regulatory interactions using network component analysis \cite{liao2003network}. We randomly divide the prior known interactions in half, and use half to learn TFA. The rest we reserve as a gold standard for validation. When we incorporate TFA and repeat our experiment combining B. subtilis datasets, we see a marked improvement in performance from using TFA. Our gains from using fused regression are preserved and even increased with the use of TFA.

@@

We present a general framework for the simultaneous inference of gene regulatory networks in multiple sources based on the introduction of constraints that penalize differences in regulatory weights thought to be similar based on prior biological knowledge. In order to solve the fused network inference problem, we develop two methods for the joint learning of gene regulatory networks, based on the use of different fusion penalties. The first method, fused L2, penalizes the squared difference of the regulatory weights of fused interactions. These constraints improve the performance of the underdetermined network inference problem, by allowing information to be shared across sources. We apply this algorithm to the problem of network inference in two distantly related biological organisms -- B. Subtilis and B. Anthracis -- and show that network recovery is improved through the introduction of fusion constraints between pairs of orthologous genes. We further demonstrate the viability of fused L2 as a method for combining data from multiple experimental platforms, where fusion is between identical pairs of genes, and as a method for the incorporation of information about operon structure, where fusion reflects the assumption that genes in the same operons have similar regulators. Because the algorithm we developed can accomodate constraints between arbitrary pairs of regulatory interactions, any biological prior representing information about expected regulatory similarity can be represented, even if the prior provides no information about the magnitude or direction of regulation. The ability to incorporate multiple data sets describing related processes, as well as multiple data types, in a principled manner, helps us take advantage of the breadth of experimentation in biology to better learn the structure of gene regulation. This approach is particularly interesting in light of the diversity of model organisms used in modern biology. Different model systems provide different advantages and disadvantages for experimental design, but without a principled mechanism for combining data from multiple sources, it is difficult to fully leverage data obtained from even a slightly different model system.

@@

Because our approach allows us to directly test -- and reject -- the hypothesis of conserved function, and because the enforcement of fusion constraints between non-conserved interactions may degrade network inference performance, we introduce an adaptive fusion method that decouples pairs of interactions for which expression data suggests a lack of conservation. Like the fused L2 approach, this method attempts to solve for networks which minimize a penalty function fitting the expression data and the fusion constraints. In the case of cross-species network inference, evolutionary links may be useful to guide the sharing of data. However, orthology may not always accurately predict functional similarity, and a fusion constraint encouraging orthologous gene, TF pairs to share similar interactions across species may not be supported by expression data. Then, when the fusion penalty is at odds with the penalty fitting expression data, there exists a tension between satisfying evolutionary constraints while accurately fitting experimental data. Our saturating penalty relaxes the fusion penalty, giving evidence that the fusion constraint is not predictive of gene interaction. We may use the analogy of a rubber band, which can become stretched when penalty functions within our objective function oppose each other; the adaptive fusion approach allows the snapping of this once the opposition is great enough. For the multiple species case, this represents orthologs which do not share similar interactions. When jointly learning networks describing processes in different cell lines, this may identify interesting context-specific behavior. Genes may be fused together on the basis of similar binding sites or chromatin features, and the relaxing of the fusion penalty indicates divergence of gene function. Adaptive fusion, therefore, is a tool for network inference as well as a method for testing network conservation.

>>>>

\title{Fused regression for multi-source gene regulatory network inference}
\author{Kari Y. Lam, Zachary M. Westrick, Christian L. M\"{u}ller, Lionel Christiaen, Richard Bonneau}
\maketitle

@@

We present two methods for network inference based on linear estimates of gene expression dynamics, extending existing linear dynamical-systems methods for network inference \cite{bonneau_predictive_2007, arrieta-ortiz_experimentally_2015}. The core of both methods is the observation that biological information about the relatedness of genes can be used to constrain which network coefficients should be similar to one another in a multi-source network inference problem (ie orthologous TFs should regulate orthologous genes), and that these constraints can be efficiently represented as penalties in a least-squares regression problem. Our first method -- fused ridge -- uses an L2 penalty on the differences between a priori similar interactions, and is useful where the relationships between networks (similarity of genes between data sets) is reliable. In the case where both networks contain an identical set of genes and TFs, this approach can be thought of as parametrically interpolating between treating the data sources separately and combining them together. However, our method allows useful pooling of data even when the overlap between genes is incomplete. Our second method -- adaptive fusion -- uses a non-concave saturating fusion penalty to simultaneously infer the constrained networks and to learn which constraints should be relaxed (ie which parts of the network are genuinely different). With this approach, we seek to test the hypothesis of conservation of interactions across networks.

@@

We test the ability of fused L2 and adaptive fusion to improve network recovery on both synthetic data and \textit{Bacillus subtilis} and \textit{Bacillus anthracis}, showing the applicability of our method in combining different datasets, and leveraging similarity across organisms as well as within a network in order to improve network inference. We explore the circumstances under which each approach is optimal, and evaluate the robustness of adaptive fusion to incorrect orthology, simulating the biologically relevant case of neofunctionalization.

@@

%We can then define $W_{i,j} = \beta_{(j-1) \times N + i, (j-1) \times M + j}$ which is the desired form.
This is equivalent to the original problem due to the block structure of matrix multiplication. In an ordinary regression problem each response variable can be solved independently as a large number of small regression problems. However, in fused regression, we append additional rows to the design matrix $A$ that link entries of $W$ associated with different response variables. As a result, these linked response variables must be solved simultaneously through vectorization. Two response variables are linked by a fusion constraint if any of the regulatory weights affecting those genes are linked by a fusion constraint. Two response variables must be solved simultaneously if there is any chain of linked response variables connecting them. However, every other response variable can be solved separately. In biological terms, the regulators of two genes (whether in the same species, or different species) must be solved together if there is a fusion constraint linking those genes' regulators, or if there is a chain of such constraints. This is important because vectorization produces a design matrix with a number of rows and columns equal to the sum of the number of rows and columns of the design matrices associated with each of the response variables (genes) being solved for. If the networks for a large number of genes are solved simultaneously, the system of equations can quickly become intractible to solve.

@@

\section{Results}
We tested the ability of fused regression to improve network inference performance, as well as the ability of adaptive fusion to learn the conservation of gene interactions, using both synthetic networks and real data. For the synthetic data, we generated random pairs of networks in which orthologous genes have similar regulatory interactions, then sampled gene expression from these networks, which we used to derive networks for comparison with the true network. For real data, we computed recovery of a known gold standard in \textit{Bacillus subtilis}. We measured network inference performance with the area under the precision recall curve (AUPR).

\subsection{Using fused regression to learn related networks}
It is known that network recovery improves with additional data \cite{bar-joseph_computational_2003}. Additional data is not always readily available; but often there is an abundance of related data, and we reason that data describing similar processes can be used to improve our ability to recover true networks. Using related data for network inference allows us to borrow statistical power from other sources, effectively increasing the sample size and boosting the sensitivity and specificity of learned interactions. We created synthetic networks that approximate two related biological processes, then evaluated performance of our fused L2 regression, which learns the networks simultaneously, sharing information about related interactions (Figure 1b). We compared recovery of the 10 TF by 200 gene networks, using fused L2 versus learning networks separately, using different amounts of simulated expression data from the two species. When the amount of data from species two is held constant, increasing the amount of data available for learning the network for species one results in a more accurate network, as expected. Using the assumption that similar genes have similar regulatory interactions, we share information about putatively conserved interactions. When we increase the amount of data from species two, we get performance gains on network one using fused L2 regression.

@@

As expected, when the conserved subgraph is very similar - when the interactions in the conserved subnetwork are nearly identical, simulating the case of closely related organisms or similar processes (Figure 2A) - increasing the weight of the fusion constraint $\lambda_S$ improved network recovery. As the size of the conserved subgraph increases, using a greater weight on the fusion penalty was beneficial.

To simulate the case of distantly related organisms, we created networks where conserved interactions are drawn from high variance distribution (Figure 2B). When the conservation is weak - when the interactions in the conserved subnetwork are nearly as different across species as randomly chosen interactions - performance may not continue to improve by increasing the fusion weight. We show that even for networks where the conserved subgraph is weakly conserved, there exists a 'sweet spot' where fusion regression improves network recovery.

\subsection{Fused regression improves performance on both the constrained and non-constrained parts of the network}
Our method learns related networks simultaneously, using predictions about conservation of TF-gene interactions to share information between sources. Our approach is useful for learning networks from similar sources such as different cell-lines from the same species, where there exists a one-to-one mapping of genes, as well as datasets where the orthology mapping does not span all the genes. This can occur when using different technology, eg microarray and RNAseq, where one method does not assay all of the genes that another one does, but the networks are expected to be very similar; or when inferring networks describing distantly related species, in which case the networks are expected to be more different than similar. When orthology is not complete, we are interested in knowing if performance gains from fused regression are limited to those interactions which have fusion constraints. To test this, we used sets of 20 TF by 200 gene synthetic networks, with varying proportions of orthologous TFs and genes. We divided networks into those interactions with fusion constraints (interactions where there exist putative analogous interactions in the second network based off of the orthology mapping between TF, gene pairs), which we call the constrained subnetwork, and interactions without fusion constraints, which we refer to as the non-constrained subnetwork. We varied the weight on the fusion penalty, lamS, and evaluated performance by computing AUPR on the constrained subnetwork, the non-constrained subnetwork, and the whole network (Figure 1C). Since the conserved subgraphs are similar to each other (more like Figure 2A than 2B), we expected performance to improve as the fusion penalty weight increased. We observed this, particularly for the constrained subnetwork. As lamS increased, interactions with fusion constraints were encouraged to be more similar; growth in performance slowed as the differences in interaction weights approach zero, and the small amount of added noise to the expression data and slight differences between the conserved networks limit the recovery from reaching completeness. Interestingly, performance gains were seen even in the portion of the network that is unconstrained by fusion. By constraining part of an undetermined system, we obtained gains in even the unconstrained part.

\subsection{Adaptive fusion successfully identifies and unfuses 'neofunctionalized' genes}
We recognize that orthology prediction is not a perfect proxy for functional conservation \cite{gabaldon_functional_2013, studer_how_2009, nehrt_testing_2011}. We implemented an adaptive fusion algorithm that attempts to optimize a nonconvex saturating penalty function on differences between fused interactions (Figure 3). Pairs of interactions which are very dissimilar even after fusion, which sit in the flat portion of this penalty function, are effectively ``unfused,'' and no further penalty is incurred as differences in interaction weights grow. Because our network inference problem is underdetermined, meaning that multiple networks would fit our data equally well, the ``unfusing'' or relaxation of the fusion penalty on constraints is convincing evidence for the rejection of a network which uses orthology to predict similarity in interaction weights. ''Unfusing'' suggests neofunctionalization of one or more of the genes involved in this pair of interactions.

We performed a simulation to assess the ability of our adaptive fusion algorithm to learn which parts of two input networks are conserved when orthology information is faulty, to mimic orthologs which are not functionally analogous  (Figure 4). We generated synthetic fused networks and introduced error in the fusion constraints by adding false positives and negatives to the orthology information given to the solver. Because we knew which entries in the orthology mapping were ``incorrect'' (not reflected in the generation of the networks), we could correctly label fusion constraints that involved one or more ``incorrect'' mappings. We verified that adaptive-fusion unfused mostly ``incorrect'' interactions, while leaving truly analogous interactions fused (Figure 4a. Note that while pairs of interactions between true orthologs are sampled randomly so that the difference in interactions tends to be small, and pairs of interactions between incorrect orthologs are sampled randomly so that the difference in interactions tends to be large, there is an element of randomness in the sampling. This results in a small fraction of interactions between incorrect othologs being similar, and similarly, a small fraction of interactions between correct orthologs being different. Consequently, some interactions between incorrect orthologs remain fused and some interactions between correct orthologs are unfused.

We compared the recovery of interaction weights which were accurately fused, and recovery of interaction weights which were inaccurately fused due to incorrect orthology information (Figure 4b). Because fused L2 heavily penalizes large differences between weights which are predicted to be similar, it is able to retrieve a more accurate network for those interactions with true fusion constraints than learning networks separately. However, the gains accomplished by sharing information through fused regression do not extend to those interactions lacking true fusion constraints, and the error remains similar to learning networks separately. Because the penalty function used in adaptive fusion resembles that of fused L2 where differences in interactions are small (Figure 3), adaptive fusion's performance on accurately fused interactions is similar to that of fused L2. Using adaptive fusion, we are able to identify which interactions should not be similar, and therefore, which orthology mappings are not predictive of functional similarity. However, because these interactions remain underdetermined, and so we are not necessarily much closer to learning their values.

\subsection{Cross-species network inference using bacterial data}
We used gene-expression data from \textit{\textit{Bacillus subtilis}} and \textit{\textit{B. anthracis}} in order to assess performance gains of fused regression on real data. Our B subtilis data set consists of 360 time-series and steady-state observations of 4891 genes during development. Our \textit{\textit{B. anthracis}} dataset consists of 72 time-series and steady-state observations of 5536 genes comprising data from developmental and iron-starvation conditions. There were 247 known Transcription factors (TFs) in the \textit{\textit{B. subtilis}} dataset, and 248 TFs in the \textit{B. anthracis} dataset.

We obtained 1870 one-to-one orthologs from Inparanoid \cite{ostlund_inparanoid_2010}, which produced 177,650 fusion-constraints between gene interactions within the two species. This number represents only $14.7\%$ of the regulatory interaction matrix in \textit{B. subtilis} and $12.9\%$ in \textit{B. anthracis}. The use of such dissimilar species is a test of whether fusion can improve network-inference even when the overlap in interactions is very small.

In order to evaluate network inference performance, and for use as priors, we used a gold standard of 2896 known \textit{B. subtilis} interactions with corresponding activation and repression sign. Of these 2896 priors, 968 had corresponding interactions in \textit{B. anthracis}. Based on our simulation results, we can expect the greatest gains in network-inference performance from fusion when the species of interest has a small number of available conditions, but data is abundant in a related species. However, in order to evaluate performance objectively a gold-standard of known interactions is necessary. As a result, we can only evaluate network recovery for \textit{B. subtilis}, and \textit{B. subtilis} also has the majority of our conditions. In order to simulate the data-poor regime, we subsampled our \textit{B. subtilis} data. Specifically, we divided our \textit{B. subtilis} data into $k$ folds, and then for each fold fit a network to the \textit{B. subtilis} data from that fold alone fused to the entire 72 \textit{B. anthracis} conditions. We can then compute performance metrics (ie AUPR) for each fold, and use their variability across folds to test for significance of any improvements relating to the use of fusion constraints.

When we applied fused L2 to our subsampled \textit{B. subtilis} and \textit{B. anthracis} datasets using optimized parameters, we did not see a significant improvement over solving the networks separately. Inspection of precision recall curves shows that there is increased precision at low values of recall. Essentially, precision increases for the high confidence part of the network but not for the low confidence part of the network. As a result, the slight improvements in performance may be more useful than would be suggested by AUPR alone. Interactions suggested by genome wide analysis are useful for target prediction, and the ranking of interactions produced by a network inference algorithms is a useful guide to what order these experiments should be carried out in. The high confidence part of the network represents interactions that are likely candidates for experimental validation, so improvement in recovery of these interactions is particularly useful in guiding research.

\subsection{Within-species fusion using similarity of promoter region}
Information about the similarity of TF-gene interactions can also come from knowledge about the promoter region; this may be a better predictor of target similarity than orthology. In bacteria, genes within the same operon are under the control of the same promoter. We predicted, therefore, that genes within the same operon will be regulated similarly by the same transcription factors. We applied fusion regression by creating fusion constraints between a given transcription factor and genes within the same operon, and showed a boost in \textit{B. subtilis} network recovery using within-species fusion (Figure 6b).

\subsection{Integrating datasets using fusion regression}
Although there are many large-scale collaborations which attempt to make protocols as uniform as possible for comparability between datasets generated by different labs \cite{kundaje_integrative_2015}, there still exists technical and biologial variability between many experiments attempting to capture the same or similar experimental condition. With the advent of new technology, such as RNAseq and single-cell RNAseq, microarray is no longer the dominant assay for genome-wide expression, but it is important to be able to combine datasets generated using different technologies. Currently, the most widely used approach to combining datasets for network inference is to learn networks from disparate datasets separately, then rank combine the networks as in Marbach et al \cite{marbach_revealing_2010}. We compared this approach to using fusion regression. We included, along with our \textit{B. subtilis} dataset, a previously published dataset containing 269 samples covering 104 conditions, obtained using a different microarray and different strain of \textit{B. subtilis} \cite{nicolas2012condition}. We compared performance when learning the networks separately, rank combining, and learning the networks simultaneously using fusion regression, and showed improvement on network inference using our fused L2 approach (Figure 6a). Importantly, our approach simultaneously learns multiple networks, rather than learning networks separately than combining into one.

\subsection{Transcription factor activity estimation integrates into fusion regression approach}
We tested a combination of our fused regression approach with a method for estimating transcription factor activities (TFA). Rather than modeling gene expression using transcription factor mRNA abundance, we attempted to fit gene expression as a function of transcription factor activity, which have been shown to be better predictors of gene expression \cite{arrieta-ortiz_experimentally_2015}. We estimated TFA based on known regulatory interactions using network component analysis \cite{liao2003network}. To test the integration of this approach with our fused regression, we asssesed the integration of \texti{B. subtilis} datasets, as in figure 6a, with the incorporation of TFA estimation. We randomly divided the prior known interactions in half, and used half to learn TFA. The rest we reserved as a gold standard for validation. As in previous studies, we observed a marked improvement in network inference when using transcription factor activity, which captures activation of transcription factor through mechanisms such as dimerization and interaction with required factors, over transcription factor expression level alone \cite{fu_reconstructing_2011}. We also obtained AUPR improvement when using fused regression on TFA, and showed that our gains from sharing information across datasets using fused regression were preserved and even enhanced by using TFA.

@@

We present a general framework for the simultaneous inference of gene regulatory networks in multiple sources based on the introduction of constraints that penalize differences in regulatory weights thought to be similar based on prior biological knowledge. In order to solve the fused network inference problem, we develop two methods for the joint learning of gene regulatory networks, based on the use of different fusion penalties. The first method, fused L2, penalizes the squared difference of the regulatory weights of fused interactions. These constraints improve the performance of the underdetermined network inference problem, by allowing information to be shared across sources. We apply this algorithm to the problem of network inference in two distantly related biological organisms -- \textit{B. subtilis} and \textit{B. anthracis} -- and show that network recovery is improved through the introduction of fusion constraints between pairs of orthologous genes. We further demonstrate the viability of fused L2 as a method for combining data from multiple experimental platforms, where fusion is between identical pairs of genes, and as a method for the incorporation of information about operon structure, where fusion reflects the assumption that genes in the same operons have similar regulators. Because the algorithm we developed can accomodate constraints between arbitrary pairs of regulatory interactions, any biological prior representing information about expected regulatory similarity can be represented, even if the prior provides no information about the magnitude or direction of regulation. The ability to incorporate multiple data sets describing related processes, as well as multiple data types, in a principled manner, helps us take advantage of the breadth of experimentation in biology to better learn the structure of gene regulation. This approach is particularly interesting in light of the diversity of model organisms used in modern biology. Different model systems provide different advantages and disadvantages for experimental design, but without a principled mechanism for combining data from multiple sources, it is difficult to fully leverage data obtained from even a slightly different model system.

@@

Because our approach allows us to directly test -- and reject -- the hypothesis of conserved function, and because the enforcement of fusion constraints between non-conserved interactions may degrade network inference performance, we introduce an adaptive fusion method that decouples pairs of interactions for which expression data suggests a lack of conservation. Like the fused L2 approach, this method attempts to solve for networks which minimize a penalty function fitting the expression data and the fusion constraints. In the case of cross-species network inference, evolutionary links may be useful to guide the sharing of data. However, orthology may not always accurately predict functional similarity, and a fusion constraint encouraging orthologous gene, TF pairs to share similar interactions across species may not be supported by expression data. Then, when the fusion penalty is at odds with the penalty fitting expression data, there exists a tension between satisfying evolutionary constraints while accurately fitting experimental data. Our saturating penalty relaxes the fusion penalty, giving evidence that the fusion constraint is not predictive of gene interaction. For the multiple species case, this represents orthologs which do not share similar interactions. When jointly learning networks describing processes in different cell lines, this may identify interesting context-specific behavior. Genes may be fused together on the basis of similar binding sites or chromatin features, and the relaxing of the fusion penalty indicates divergence of gene function. Adaptive fusion, therefore, is a tool for network inference as well as a method for testing network conservation.
